### 2014-10-25 motivation.tex
M

\section{Motivation}

In this section, we motivate the need for declarative reasoning for eventually
consistent data store.

\subsection{System Model}

We assume a system model where the eventually consistent database is organized
as a collection of key-indexed, object-valued tables. The objects stored in the
tables are instances of the same replicated data type. The database is composed
of a number of \emph{replicas}, and each object is fully replicated across all
the replicas. Under this model, a client request for an operation with no
application-level consistency requirement can be serviced if at least one of
the replicas is reachable. Thus, with the addition of new replicas, and placing
them close to the clients, leads to improved throughput and reduced latencies.

@@

Suppose our goal is to implement a bank account service on top an eventually
consistent data store, with the following schema,

\begin{codesql}
Create Table BankAccount (
  userID int PRIMARY KEY,
  userName string UNIQUE,
  balance float CHECK (balance >= 0))
\end{codesql}

\noindent with the following \emph{operations}.

\begin{codehaskell}
type UserName = String
-- returns false if the user name is already taken
addAccount     :: UserName -> Bool
getBalanceName :: UserName -> Float
depositName    :: UserName -> Float -> ()
-- returns false if the account has insufficient balance
withdrawName   :: UserName -> Float -> Bool
\end{codehaskell}

Notice that the operations manipulating the table take user name as input.
Since eventually consistent databases are non-relational~\cite{}, we need to
maintain a \emph{secondary index} to look up records by user name.

\begin{codesql}
Create Table UserIndex (
  userName string PRIMARY KEY,
  userID int
  FOREIGN KEY (userID)
    REFERENCES BankAccount (userID) )
\end{codesql}

\begin{figure}[t]
\centering
\subfigure[Negative Balance]{\label{fig:negativeBalanceAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation2}}
\hfill
\subfigure[Missing update]{\label{fig:missingUpdateAnomaly}\includegraphics[width=0.26\columnwidth]{Figures/Motivation1}}
\hfill
\subfigure[Monotonicity violation]{\label{fig:monotonicityAnomaly}\includegraphics[width=0.36\columnwidth]{Figures/Motivation3}}
\caption{Anomalies possible under eventual consistency for the get balance operation.}
\label{fig:cleanliness_examples}
\end{figure}

@@

In a traditional database system, the programmer need only to define the
application-level integrity constraints along with the schema, and the database
management system automatically enforces the necessary coordination. However,
under eventual consistency, the onus is on the programmer to ensure not only
that the integrity constraints are preserved, but also to prevent any
inconsistencies being exposed to the user. For example, it is easy to see that
the withdraw operations must be strongly consistent operations to preserve the
integrity constraint. However, it is possible that the get balance operation
still returns a negative balance.

Figure~\ref{fig:negativeBalanceAnomaly} illustrates such a potential execution.
Assume that all operations are on the same object, and the balance on the
account was initially 0. Here, \cf{vis} edges capture the visibility
relation between the operations. In this case, the withdraw sees the deposit,
but the get balance operation only witnesses the withdraw. Such an execution is
possible under eventual consistency, where the \emph{effect} (update)
corresponding to the withdraw reaches the replica to which the get balance is
applied to before the causally preceding deposit effect. Alternatively, it is
possible that a get balance operation does not see the effects from its own
session (Figure~\ref{fig:missingUpdateAnomaly}), leading the user to
incorrectly conclude that the previous update failed to go through. Here, the
\cf{so} edge represents the session order that exists between the actions
on the same thread.

The anomalies discussed so far arise from weakly consistent behaviors on the
same object. With operations on multiple objects on distinct tables, the
programmer on eventually consistent databases has to take care of preserving
integrity constraints for foreign keys, materialized views and secondary
indexes. For example, Figure~\ref{fig:monotonicityAnomaly} illustrates the
anomaly that arises from the lack of monotonic update visibility. Adding a new
account involves a conditional insertion into the index table, followed by an
insertion into the bank account table. Many eventually consistent data stores
propose write-only transactions~\cite{}, and let us assume that add user
operation is marked as a write-only transaction. Despite this, under eventual
consistency, it is possible for a \cf{getBalanceName} operation to witness the
first insertion but not the second, exposing an inconsistent state. The two
reads may be served by different replicas, where the only the former has
witnessed the effects of add user transaction. Assigning the correct store
consistency level for the operations in this example such that the
application-level constraints are preserved is a non-trivial task.

@@

Quelea automates this error-prone task; the programmer need only to express the
application-level consistency constraints in the contract language. Replicated
data types in Quelea are defined as reductions over the set of effects similar
to operation-based CRDTs~\cite{}. The following data type definitions capture
the operations and effects on the bank account and user index objects.

\begin{codehaskell}
data BankAcc = AddUser UserName | Deposit Float
             | Withdraw Float | GetBalance
data UserIdx = AddName UserID | GetName
\end{codehaskell}

Recall that the key of the bank account table is user id, which maps to a
particular bank account object. In Quelea, an object state is nothing but the
set of effects (called the \emph{context}) on this object. As such,
\cf{Deposit} and \cf{Withdraw} effects simply include the amount deposited or
withdrawn, and \cf{AddUser} effect includes the user name. Similarly, on the
user index table, where the key is user name, the value is simply an effect
(\cf{AddName}) binding some user id at this key. \cf{GetBalance} and
\cf{GetName} correspond to the corresponding read-only operations. The complete
definition of the operations on bank account and user index objects is given
below.

>>>>

\section{Motivation}
\label{sec:motivation}

In this section, with help of a bank account example, we motivate the
need for a framework to express and enforce fine-grained
application-level consistency requirements under weak consistency
models. 
% \GK{The following text needs to be polished}

% We assume a system model where the eventually consistent database is organized
% as a collection of key-indexed, object-valued tables. The objects stored in the
% tables are instances of the same replicated data type. The database is composed
% of a number of \emph{replicas}, and each object is fully replicated across all
% the replicas. Under this model, a client request for an operation with no
% application-level consistency requirement can be serviced if at least one of
% the replicas is reachable. Thus, with the addition of new replicas, and placing
% them close to the clients, leads to improved throughput and reduced latencies.

@@

Suppose our goal is to implement a highly available bank account
service on top of an eventually consistent data store. For the sake of
presentation, we restrict ourselves to bank account objects (records)
with following simple schema:

\begin{codesql}
Create Table BankAccount (
  accountNo int PRIMARY KEY,
  balance float CHECK (balance >= 0))
\end{codesql}

\noindent and with support for the following \emph{operations}:

\begin{codehaskell}
type AccountNo = Int
type Amount = Float
getBalance :: AccountNo -> Amount
deposit    :: AccountNo -> Amount -> ()
-- returns false if the account has insufficient balance
withdraw   :: AccountNo -> Amount -> Bool
\end{codehaskell}

\begin{figure}[t]
\centering
\subfigure[Unsafe Withdraw]{\label{fig:unsafeWithdrawAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation4}}
\hfill
\subfigure[Negative Balance]{\label{fig:negativeBalanceAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation2}}
\hfill
\subfigure[Missing update]{\label{fig:missingUpdateAnomaly}\includegraphics[width=0.26\columnwidth]{Figures/Motivation1}}
%\hfill
%\subfigure[Monotonicity violation]{\label{fig:monotonicityAnomaly}\includegraphics[width=0.36\columnwidth]{Figures/Motivation3}}
\caption{Anomalies possible under eventual consistency for the get balance operation.}
\label{fig:cleanliness_examples}
\end{figure}

The sole attribute of a \cf{BankAccount} object, identified by its unique
\cf{accountNo}, is its current \cf{balance}. A \cf{deposit} operation
on a \cf{BankAccount} adds some \cf{Amount} to its \cf{balance},
whereas a \cf{withdraw} operation subtracts. Operation \cf{getBalance}
returns the current \cf{balance} in the account. A \cf{withdraw}
operation succeeds and returns \cf{True} only if there is sufficient
balance in the account. Otherwise, it fails and returns \cf{False}.

If we were to use a traditional database system to store
\cf{BankAccount} objects, all that needs to be done is to define the
\cf{BankAccount} schema, as shown above. The database management
system effectively sequentializes all accesses to the database, and
automatically enforces the required integerity constraints on the
data, so that it is never possible for a \cf{getBalance} operation to
return a negative balance, or for a \cf{withdraw} to succeed without
having sufficient balance in the account. However, under eventual
consistency, such anomalies are bound to occur as demonstrated in the
following subsection.

@@

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly}.
Assume that all operations in the figure are on the same
\cf{BankAccount} object. Execution contains two concurrent client
sessions, where one of the sessions executes a \cf{withdraw} followed
by a \cf{getBalance} operation. We say that \c{withdraw} precedes
\cf{getBalance} in \emph{session order}, and visualize it as a
directed \cf{so} edge from the former to the later. Edges labelled
\cf{vis} in the figure capture the \emph{visibility} relation between
operations; an operation is said to be \emph{visible} to other
operation if the later can witness the effects of the former. Two
operations that are not related either via \cf{vis} or \cf{so} are
said to be concurrent. The figure shows an execution where concurrent
\cf{withdraw}s in different sessions succeed independently, but a
subsequent \cf{getBalance} operation that witnesses both the
\cf{withdraw}s reports a negative balance.

It is easy to see that the only way to ensure that integrity
constraint on \cf{balance} is to prevent concurrent withdrawls on the
\cf{BankAccount}. This can be done by insisting that \cf{withdraw} be
executed with strong consistency. Although this prevents unsafe
withdrawls from going through, a strongly consistent \cf{withdraw}
itself is not sufficient to prevent \cf{getBalance} from ever
reporting a negative balance to the end user.

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly},
which consists of three concurrent sessions performing a \cf{deposit},
a \cf{withdraw}, and a \cf{getBalance}, respectively, on the same
\cf{BankAccount}. Assume that initial balance in the account is zero.
As the \cf{vis} edge indicates, operation \cf{withdraw(50)} in session
2, can witnesses the effects of \cf{deposit(100)} from session 1,
leading it to rightfully conclude that the current balance is 100.
Consequently, \cf{withdraw(50)} is performed safely. The effect of
this \cf{withdraw} operation subsequently becomes visible to the
\cf{getBalance} in Session 3. However, \cf{getBalance}, which can only
see the effect of \cf{withdraw} from session 2, but not \cf{deposit}
from session 1, reports the balance of negative 50 to the user.

Negative balance is not the only inconsistency that the user is
exposed to under an eventually consistent bank account. Figure
~\ref{fig:missingUpdateAnomaly} shows an execution where
\cf{getBalance} operation in a session does not witness the effects of
\cf{withdraw} operation performed in the same session, possibly
because \cf{getBalance} was served by a replica that hasn't yet merged
the \cf{withdraw} effect, which could have been served by a different
replica. This anomaly leads the user to incorrectly conclude that the
\cf{withdraw} operation failed to go through.

Although it is easy to understand the reasons behind the occurance of
the aforementioned negative balance and missing update anomalies, the
fix is not quite apparent. Making \cf{getBalance} a strongly
consistent operation is definitely sufficient to avert anomalies, but
is it really necessary? Given that a strongly consistent operation is
significantly expensive, both in terms of latency and monetary value, we ideally
would not want to request strong consistency, unless there are
no alternatives available. Unfortunately, this requires us to
understand subtle differences in semantics of various kinds of weak
consistency alternatives, such as timeline consistency and session
guarantees, offered by off-the-shelf eventually consistent data
stores. This process is error-prone, and becomes further complicated
in presence of transactions, as demonstrated in Sec.
~\ref{sec:transactions}.

@@

\name automates this error-prone task by letting the programmer
declare application-level consistency constraints as \emph{contracts},
obviating the need to understand store-level consistency guarantees,
and relieving programmer from having to match application
requirements with appropriate store-level alternatives. The
cornerstone of declarative reasoning in \name is the programming model
built around operation-based CRDTs~\cite{shapiroCRDT}. A replicated
data type in \name is defined as a tagged union type, where each tag
is the name of a unique operation defined on the data type. 
%types in Quelea are defined as reductions over the set of effects.
%The following data type definitions capture the operations and effects
%on the bank account and user index objects.
For example, following is the definition of the \cf{BankAccount}
data type in \name:
\begin{codehaskell}
data BankAccount =  Deposit Float | Withdraw Float 
                 | GetBalance
\end{codehaskell}

In Quelea, an object state is nothing but the set of effects (called
the \emph{context}) on this object. As such, \cf{Deposit} and
\cf{Withdraw} effects simply include the amount deposited or
withdrawn, and \cf{AddUser} effect includes the user name. Similarly,
on the user index table, where the key is user name, the value is
simply an effect (\cf{AddName}) binding some user id at this key.
\cf{GetBalance} and \cf{GetName} correspond to the corresponding
read-only operations. The complete definition of the operations on
bank account and user index objects is given below.
